---
title: "Sanitizing and cleaning the Oceania-UK dataset"
output:
  html_document:
    keep_md: yes
---

## Introduction

Here we apply a sequence of steps to reproducibly sanitize the oceania-uk dataset. We start from `oceania-uk-data.csv`, which is the result of applying the following steps to the original spreadsheet:

1. Insert column _continent_ to Australia data table between _population size_ and _life expectancy_.
2. Change column A column header from _continent_ to _country_. (This is for the Australia data table.)
3. Move the other (than Australia) per-country tables (only UK at present) under the Australia table.

No additional manipulation has been done yet.

## Loading libraries and other setup

```{r}
# the name of the file containing the dataset:
datafile <- "oceania-uk-data.csv"

# the name of the metadata file:
metafile <- paste(paste(strsplit(datafile, split = "-")[[1]][c(1,2)],
                        collapse="-"),
                  "metadata.txt",
                  sep = "-")
metafile

# the name of the metadata file:
outfile <- paste(paste(strsplit(datafile, split = "-")[[1]][c(1,2)],
                       collapse="-"),
                 "sanitized.csv",
                 sep = "-")
outfile
```

## Moving metadata out into a separate file

The first two lines are metadata. Read those in and write out to a metadata file:

```{r}
file.header <- scan(datafile, 
                    what = "character", 
                    sep = ",", 
                    nlines = 2)
file.header

writeLines(file.header[1], metafile) # We only want what is in the first cell
```

## Sanitizing the data

Read in data, standardizing NA values, skipping blank lines, properly setting column header names:

```{r}
data.in <- read.table(datafile, 
                      sep = ",", 
                      skip = 4, 
                      col.names = c("country", 
                                    "year", 
                                    "pop", 
                                    "continent", 
                                    "lifeExp", 
                                    "gdpPercap", 
                                    "blank", 
                                    "Notes"), 
                      blank.lines.skip=TRUE, 
                      na.strings = c("N/A", "NA", ""))
```

Remove the empty column:
```{r}
data.in <- subset(data.in, select = -c(blank))
```

Fix the typo in the country column and remove excess factor levels:
```{r}
data.in$country[data.in$country == "Australa"] <- "Australia"
data.in$country <- factor(data.in$country)

# Test: we should be left with 2 factors now in country:
if (nlevels(data.in$country) > 2) {
  cat("Data integrity alert: more than 2 factors for country")
}
```

Fix the typo in the population column:
```{r}
pop.is.typo <- is.na(as.numeric(as.character(data.in$pop)))
pop.typo <- strsplit(as.character(data.in$pop[pop.is.typo]),"")[[1]]
pop.typo[pop.typo == "O"] <- "0"
data.in$pop <- as.numeric(as.character(data.in$pop))
data.in$pop[pop.is.typo] <- as.numeric(paste(pop.typo,collapse=""))
```

## Write sanitized data to csv

```{r}
write.csv(data.in, file = outfile)
```
